---
output:
  github_document:
    toc: false
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Avoiding Loops
An introductory tutotrial on possible means to avoid loops altogether in **R**

This tutorial introduces you some exciting alternatives to classic loops in **R**. While classic `for()` and `while()` loops are quite faster now in **R**, they still suffer considerable time lag when compared to loop structures in other high end languages (**C++** for example). They actually happen to most of the time not be the best option when a task is to be repeated several times on a large number of datasets. This is due to no fault of their own but the fact that **R** isn't a vectorized language.  The tutorial adopts a hands-on approach to demonstrate the offerings made available via existing functions (**apply()**, **sapply()**, **lapply()**, **tapply()** and **mclapply**) and packages (**purrr**, **repurrrsive** and **futures**). The tutorial ends with an introduction to parallel programming and some packages that can help achieve this essential modern day computing technique in **R**.

##  Installation of Nedded Packages

```{r}
#install.packages(tidyverse)
#install.packages(repurrsive)
```

## R Session Info

```{r}
sessionInfo()
```

##  Contributors

- Owokotomo Olajumoke Evangelina
- Sengupta Rudravdev

##  Useful links

[![Travis-CI Build Status](https://github.com/jennybc/repurrrsive)](https://github.com/jennybc/repurrrsive)

[![Travis-CI Build Status](https://purrr.tidyverse.org/)](https://purrr.tidyverse.org/)

[![Travis-CI Build Status](https://r4ds.had.co.nz/)](https://r4ds.had.co.nz/)
